---
title: "Getting Started"
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

## What the package installs

`gdalraster.windows` downloads a verified GDAL build runtime, compiles
[`gdalraster`](https://firelab.github.io/gdalraster/) against it, and vendors
the complete native dependency closure into the installed `gdalraster`
package. The result carries:

- GDAL and dependency DLLs under `libs/x64`;
- matching GDAL and PROJ data under `gdal/` and `proj/`;
- pure-Python `osgeo_utils` under `python/`; and
- a provenance manifest recording the bundle used for the build.

No `.Rprofile` hook, `PATH` edit, DLL preload, or GDAL/PROJ environment
variable is needed in later sessions.

## Prerequisites

- Windows and R.
- The matching [Rtools](https://cran.r-project.org/bin/windows/Rtools/)
  release. Compilation is the only step the prebuilt GDAL bundle cannot
  remove.

## One-time setup

Install the package, then build into your regular user library:

```{r}
#| eval: false
pak::pak("jimbrig/gdalraster.windows")
gdalraster.windows::gdal_setup()
```

`gdal_setup()` prints one plan, installs or skips the runtime idempotently,
builds and vendors `gdalraster`, provisions embedded Python when a system
CPython is available, and verifies the result in fresh processes.

Interactive sessions ask before replacing an existing `gdalraster` in the
destination library; non-interactive replacement requires `force = TRUE`.

For a non-destructive isolated library:

```{r}
#| eval: false
result <- gdalraster.windows::gdal_setup(isolated = TRUE)
result$lib
```

## Everyday use

After setup, `gdalraster.windows` does not need to be attached:

```{r}
#| eval: false
library(gdalraster)
gdalraster::gdal_global_reg_names()
```

For an isolated install, load from that library:

```{r}
#| eval: false
library(
  gdalraster,
  lib.loc = file.path(
    tools::R_user_dir("gdalraster.windows", "data"),
    "library"
  )
)
```

## Status and verification

```{r}
#| eval: false
gdalraster.windows::gdal_sitrep()
gdalraster.windows::gdal_verify()
```

`gdal_verify()` uses fresh `Rscript --vanilla` processes. It checks the
Algorithm API, GEOS and CRS support, Arrow, Parquet, HDF5, and netCDF driver
registration, a first Parquet open, and the GeoPackage Python validator when
Python provisioning is ready.

## Updating

```{r}
#| eval: false
gdalraster.windows::gdal_update()
```

The runtime and built package manifests let this operation distinguish a
current build from one that must be rebuilt.

## Offline setup

Download a bundle release asset on a connected machine, transfer it, then use:

```{r}
#| eval: false
gdalraster.windows::gdal_setup(
  local_zip = "C:/Downloads/gdal-bundle-v3.13.2-windows-x64.zip"
)
```

See the [Runtime Guide](runtime-guide.html) for layout and provenance details,
[Architecture](architecture.html) for the loading model, and
[Troubleshooting](troubleshooting.html) for diagnosis.
