---
title: "Getting Started"
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
#| eval: false
library(gdalraster.windows)
```

## What this package does

`gdalraster.windows` provides a reliable Windows workflow for using [`gdalraster`](https://firelab.github.io/gdalraster/) against a modern, self-contained [GDAL](https://gdal.org/) runtime built in CI — including features the default [Rtools](https://cran.r-project.org/bin/windows/Rtools/) GDAL lacks (notably the [GDAL Algorithm API](https://gdal.org/en/stable/api/gdalalg_cpp.html) with [`muparser`](https://beltoforion.de/en/muparser/), plus [Arrow](https://arrow.apache.org/)/[Parquet](https://parquet.apache.org/), [HDF5](https://www.hdfgroup.org/solutions/hdf5/), [NetCDF](https://www.unidata.ucar.edu/software/netcdf/), and [SpatiaLite](https://www.gaia-gis.it/fossil/libspatialite/index) drivers).

It does three things:

1. **Installs a GDAL runtime bundle** — a prebuilt, verified-closed set of DLLs, runtime data, and pure-python utilities published as a [GitHub release asset](https://github.com/jimbrig/gdalraster.windows/releases).
2. **Builds `gdalraster` from source** against that bundle, into an isolated library so nothing you already have installed is touched.
3. **Activates the runtime per session** so the Windows loader, GDAL, and PROJ all resolve against the bundle.

Everything is non-destructive by default: the runtime lives under a package-managed user data directory, and the source-built `gdalraster` goes into its own library path rather than replacing a CRAN install.

## Quick start

One-time setup (using [pak](https://pak.r-lib.org/)):

```{r}
#| eval: false
pak::pak("jimbrig/gdalraster.windows")

# download and install the GDAL runtime bundle (latest release asset)
gdalraster.windows::install_gdal_runtime()

# build gdalraster from source against that runtime
gdalraster.windows::install_gdalraster()
```

Then load and verify:

```{r}
#| eval: false
gdalraster.windows::load_gdalraster()

gdalraster::gdal_global_reg_names()
#> [1] "raster" "vector" "convert" "info" ... (non-empty)
```

A non-empty algorithm registry is the key success signal — on a broken toolchain/runtime combination it comes back as `character(0)`.

## Everyday use

After setup, attaching the package is enough. Its load hook automatically activates the installed runtime and prepends the isolated `gdalraster` library to `.libPaths()`:

```{r}
#| eval: false
library(gdalraster.windows)
library(gdalraster)

gdalraster::gdal_global_reg_names()
```

If you prefer explicit control over the load order:

```{r}
#| eval: false
gdalraster.windows::load_gdal_dll()

library(gdalraster)
gdalraster::gdal_global_reg_names()
```

To make sessions bootstrap without attaching this package at all, install a managed `.Rprofile` hook (optional, and clearly delimited in the file):

```{r}
#| eval: false
gdalraster.windows::add_gdal_rprofile_hook()
```

## Verifying the setup

`verify_gdalraster_runtime()` runs the full chain — activate runtime, load `gdalraster`, check the algorithm registry — and reports what it finds:

```{r}
#| eval: false
gdalraster.windows::verify_gdalraster_runtime()
#> v gdalraster is ready ("3.13.1")
```

## Offline / air-gapped machines

`install_gdal_runtime()` downloads from GitHub Releases by default. Without network access, download the release asset from <https://github.com/jimbrig/gdalraster.windows/releases> on a connected machine, transfer it, and install directly:

```{r}
#| eval: false
gdalraster.windows::install_gdal_runtime(
  local_zip = "C:/Downloads/gdal-ucrt64-v3.13.1-windows-x64.zip"
)
```

## Where things are installed

| Component | Default location | Override |
|-----------|------------------|----------|
| GDAL runtime | `tools::R_user_dir("gdalraster.windows", "data")` | `configure_gdal_home()`, `options(gdalraster.windows.gdal_home = ...)`, or the `GDALRASTER_WINDOWS_GDAL_HOME` environment variable |
| Source-built `gdalraster` | isolated `library/` folder under the same user data directory | `install_gdalraster(lib = ...)` |

To install the source build into your regular user library instead — so plain `library(gdalraster)` resolves it without this package's helpers — pass the library explicitly:

```{r}
#| eval: false
gdalraster.windows::install_gdalraster(lib = .libPaths()[1])
```

This replaces any existing `gdalraster` (e.g. the CRAN binary) in that library, and the bundled runtime must still be activated before the package loads in each session.

## Next steps

- The [Runtime Guide](runtime-guide.html) covers session activation, configuration options, the startup hook, and embedded-python algorithms in depth.
- [Architecture](architecture.html) explains why this package exists and how the runtime bundle is built.
- [Troubleshooting](troubleshooting.html) walks through the common failure modes and their fixes.
