---
title: "Runtime Guide"
vignette: >
  %\VignetteIndexEntry{Runtime Guide}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

## Runtime and installed-package roles

The managed GDAL runtime is a build-time SDK. It provides headers, import
libraries, DLLs, data, and Python utilities to
`gdal_build_gdalraster()`. Runtime use is served entirely by files vendored
inside the resulting `gdalraster` package.

```text
managed gdal_home/
  bin/*.dll
  include/
  lib/
  share/gdal/
  share/proj/
  python/osgeo_utils/
  MANIFEST.dcf

installed gdalraster/
  libs/x64/gdalraster.dll
  libs/x64/libgdal-*.dll
  libs/x64/<dependency DLLs>
  gdal/
  proj/
  python/osgeo_utils/
  gdalraster.windows-build.dcf
```

`library.dynam()` loads `gdalraster.dll` with the package's `libs/x64`
directory as its DLL path. Windows resolves the dependent `libgdal` and its
dependency closure from that same directory, including when another GDAL is
earlier on `PATH`.

## Runtime location

`gdal_home()` resolves:

1. `options(gdalraster.windows.gdal_home = "...")`;
2. `GDALRASTER_WINDOWS_GDAL_HOME`; then
3. the package-managed user data directory.

The option and environment variable are useful for tests, CI, and offline
SDKs. They do not affect later loading of a vendored `gdalraster`.

## Plumbing API

Use the lower-level functions when setup must be split:

```{r}
#| eval: false
gdalraster.windows::gdal_install_runtime()
gdalraster.windows::gdal_build_gdalraster()
gdalraster.windows::gdal_enable_python()
gdalraster.windows::gdal_verify()
```

`gdal_install_runtime()` is idempotent. A matching `Bundle-Tag` in
`MANIFEST.dcf` is skipped. A newer release is reported without destructive
replacement; `gdal_update()` or `force = TRUE` performs replacement.

`gdal_build_gdalraster()` stages a complete package before replacing an
existing installation. If the destination is locked, the old package remains
unchanged.

## Provenance and staleness

The runtime manifest records:

- bundle tag and GDAL version;
- release asset name;
- installation time; and
- installer package version.

The build manifest records the bundle tag and GDAL version plus R, platform,
compiler, build time, and builder package version. `gdal_sitrep()` compares the
two tags. A mismatch means `gdalraster` must be rebuilt.

## Embedded Python

Some algorithms, including `gdal driver gpkg validate`, embed CPython and
import `osgeo_utils`. `gdal_enable_python()` writes a managed file named
`gdalraster-windows-osgeo-utils.pth` into system CPython's `site-packages`.
The file points at the vendored `gdalraster/python` directory.

This design does not set `PYTHONPATH`, does not modify user or machine
environment variables, and does not affect isolated virtual environments.
When no system CPython is available, core `gdalraster` remains usable and the
Python-specific check is reported as unavailable.

## Updates

```{r}
#| eval: false
gdalraster.windows::gdal_update()
```

An update replaces the SDK, rebuilds from source, re-vendors every runtime
component, re-points the managed `.pth`, and verifies in a fresh process.

For an isolated installation:

```{r}
#| eval: false
gdalraster.windows::gdal_update(isolated = TRUE)
```

## Uninstall

```{r}
#| eval: false
gdalraster.windows::gdal_uninstall(force = TRUE)
```

The uninstall removes the selected package from `.libPaths()[1]`, the SDK, and
the managed Python path file. Pass `isolated = TRUE` when the managed build
was placed in the isolated library.
