---
title: "Troubleshooting"
vignette: >
  %\VignetteIndexEntry{Troubleshooting}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
#| eval: false
library(gdalraster.windows)
```

## Baseline triage

Run these in order; most problems surface at one of these steps.

```{r}
#| eval: false
# 1) where is the active runtime, and is the DLL there?
gdalraster.windows::gdal_home()
list.files(
  file.path(gdalraster.windows::gdal_home(), "bin"),
  pattern = "^libgdal-.*\\.dll$"
)

# 2) activate runtime (PATH, data vars, PYTHONPATH, DLL preload)
gdalraster.windows::activate_gdal_runtime()

# 3) end-to-end verification
gdalraster.windows::verify_gdalraster_runtime()
```

If all three pass, the runtime contract is intact and the problem is elsewhere.

## `gdal_global_reg_names()` returns `character(0)`

Most likely the wrong GDAL DLL was resolved first, or activation never ran in this session.

- Run `gdalraster.windows::activate_gdal_runtime()` in a **fresh** R session, then load `gdalraster`.
- Inspect `Sys.getenv("PATH")` for competing GDAL installations (Rtools, conda/pixi, OSGeo4W) appearing before the bundle's `bin/`.
- Confirm `gdalraster` was built by `install_gdalraster()` against the bundle, not installed as the CRAN binary (which statically links Rtools' GDAL).

## `LoadLibrary failure` when loading gdalraster

The Windows loader could not resolve `libgdal-*.dll` or one of its transitive dependencies. `activate_gdal_runtime()` fails loudly when the preload of `libgdal-*.dll` itself fails, so a `LoadLibrary failure` surfacing only at `library(gdalraster)` usually means the runtime was never activated in this session.

- Activate the runtime before `library(gdalraster)` — or use `gdalraster.windows::load_gdalraster()`, which does both.
- Isolate profile effects with a clean subprocess:

```powershell
Rscript -e "gdalraster.windows::activate_gdal_runtime(); library(gdalraster); print(length(gdalraster::gdal_global_reg_names()))"
```

- Inspect the dependency tree from an [MSYS2](https://www.msys2.org/)/[Rtools](https://cran.r-project.org/bin/windows/Rtools/) shell with [ntldd](https://github.com/LRN/ntldd): `ntldd -R <gdal_home>/bin/libgdal-*.dll` — any entry that resolves outside the bundle and outside Windows system paths indicates a missing bundled DLL. Or list direct imports with `objdump -x <gdal_home>/bin/libgdal-*.dll | findstr "DLL Name"`.

A DLL can be "missing" even when CI verified the bundle: GitHub runner images ship products that install DLLs into `System32` (e.g. the Microsoft ODBC Driver for SQL Server, `msodbcsql17.dll`), so a dependency on them resolves in CI but not on end-user machines. The build disables such linkage (`GDAL_USE_MSSQL_ODBC=OFF`) and bundle verification rejects known non-OS `System32` DLLs, but any import in the `objdump` output that is neither in the bundle's `bin/` nor a Windows system DLL is the culprit.

## `LoadLibrary failure: A dynamic link library (DLL) initialization routine failed`

This is a different failure class from "module could not be found" (Windows error 1114 vs 126). Every dependency was *found* and mapped into the process, but one of them failed its own initialization (`DllMain` / C++ static constructors), which makes the whole `libgdal-*.dll` load fail. Two properties make it confusing:

- **Dependency-closure checks cannot detect it.** The bundle can be verified complete (`ntldd` reports every import resolved) and still fail this way — the broken DLL is present, it just cannot initialize. This is why bundle builds are also *load-tested* in CI (every DLL is passed through `LoadLibrary` in a plain process before an asset is published).
- **The same DLL bytes can load in one process and fail in another**, because its initializer's behavior depends on which copies of *its* dependencies the loader resolved (first-loaded-wins by base name).

To isolate the culprit, sweep-load the bundle's DLLs individually in a fresh session — the failing one is rarely `libgdal` itself:

```{r}
#| eval: false
bin <- file.path(gdalraster.windows::gdal_home(), "bin")
withr::local_envvar(PATH = paste(bin, Sys.getenv("PATH"), sep = ";"))
for (dll in list.files(bin, pattern = "\\.dll$", full.names = TRUE)) {
  ok <- tryCatch(
    { dyn.load(dll, local = FALSE, now = TRUE); TRUE },
    error = function(e) FALSE
  )
  if (!ok) cat("FAILS TO INITIALIZE:", basename(dll), "\n")
}
```

Historical instance: bundles published before July 2026 shipped `libpodofo.dll` (a PDF-driver backend), whose static initializers run OpenSSL setup during `DllMain` and fail against the MSYS2-built `libcrypto-3-x64.dll` — in any process, R or not. Current bundles exclude the PDF driver entirely (`GDAL_ENABLE_DRIVER_PDF=OFF`) and CI refuses to publish a bundle containing any DLL that fails to load. If you see this error with a current bundle, reinstall the runtime first (`install_gdal_runtime(overwrite = TRUE)`), then run the sweep above and report the failing DLL in an issue.

## `install_gdal_runtime(overwrite = TRUE)` cannot delete the old runtime

Deleting `gdal_home` fails when its DLLs are mapped into a running process — including the current session: the auto-bootstrap preloads the runtime at package load and prepends its `bin` to `PATH`, so DLLs loaded later in the session resolve dependencies from the runtime directory by module name (the release download itself loads the curl package, which maps the runtime's `zlib1.dll`). Those mappings live outside R's DLL registry, so unloading `libgdal` cannot release them. The installer handles this case itself: mapped DLLs cannot be deleted but can be renamed, so leftovers this session still maps are moved into a stale sibling directory (`<gdal_home>.stale-<pid>-<timestamp>`) and the install proceeds; later installs delete stale directories once their locks are gone. When the installer reports this, restart R before building or loading `gdalraster` so the new runtime is used.

The installer refuses to run at all while `gdalraster` is loaded (its DLL pins the runtime — restart R first). If the install still aborts with a lock error, another process holds locks without delete sharing: other R sessions using the runtime, or File Explorer windows/preview panes open on the runtime directory. Close them ([PowerToys File Locksmith](https://learn.microsoft.com/en-us/windows/powertoys/file-locksmith) or Resource Monitor identify the locker) and rerun.

## `ModuleNotFoundError: No module named 'osgeo_utils'`

Raised by embedded-python algorithms such as [`gdal driver gpkg validate`](https://gdal.org/en/stable/programs/gdal_driver_gpkg_validate.html). See the [Runtime Guide](runtime-guide.html) for how the embedded-python layer works.

- Check the bundle has the python layer:

```{r}
#| eval: false
dir.exists(file.path(gdalraster.windows::gdal_home(), "python", "osgeo_utils"))
```

- If missing, the installed runtime predates python support — reinstall:

```{r}
#| eval: false
gdalraster.windows::install_gdal_runtime(overwrite = TRUE)
```

- Re-activate and confirm `Sys.getenv("PYTHONPATH")` contains `<gdal_home>/python`.
- A `python.exe` must be discoverable on `PATH` for GDAL to embed an interpreter at all (the GDAL debug stream shows which python/libpython it loads).

Note: `validate_gpkg` treats the compiled `osgeo` bindings as optional; the bundle intentionally omits them, which skips only tiled gridded coverage content checks.

## CRS / projection errors despite GDAL loading

`GDAL_DATA` / `PROJ_DATA` are not pointing at the bundle's `share/` tree. `activate_gdal_runtime()` sets all three (`GDAL_DATA`, `PROJ_LIB`, `PROJ_DATA`); check whether something later in your startup overwrites them.

## Download failures during `install_gdal_runtime()`

On machines without network access (or behind strict proxies), install from a manually transferred release asset:

```{r}
#| eval: false
gdalraster.windows::install_gdal_runtime(
  local_zip = "C:/Downloads/gdal-ucrt64-v3.13.1-windows-x64.zip"
)
```

Assets are published at <https://github.com/jimbrig/gdalraster.windows/releases>.

## `install_gdalraster()` reports "source install did not produce an installed package"

If `install_gdalraster()` fails with this error and R also printed a warning like `package '…tar.gz' is not available for this version of R`, the underlying cause was that `install.packages()` received `repos` set to a CRAN mirror together with a local file path. In that combination R looks the path up as a *package name* in the repository rather than installing from the file, so the install silently does nothing.

This was fixed in `gdalraster.windows` ≥ 0.2.1 by always calling `install.packages()` with `repos = NULL` for local-file installation. If you encounter this with an older version, upgrade the package:

```{r}
#| eval: false
pak::pak("jimbrig/gdalraster.windows")
```

## After a GDAL version upgrade

- Reinstall the runtime (`install_gdal_runtime(overwrite = TRUE)`), then rebuild `gdalraster` against it (`install_gdalraster()`) — the previous `gdalraster.dll` is bound to the previous GDAL's ABI/import names.
- Never assume the GDAL SONAME is stable across versions; everything in this package discovers `libgdal-*.dll` by glob, and your own scripts should too.
