---
title: "Troubleshooting"
vignette: >
  %\VignetteIndexEntry{Troubleshooting}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

## Start with the sitrep

```{r}
#| eval: false
gdalraster.windows::gdal_sitrep()
gdalraster.windows::gdal_verify()
```

The default inspects `.libPaths()[1]`. Pass `lib` for a custom library, or
`isolated = TRUE` for the package-managed isolated library. The sitrep
reports the GDAL runtime tag, build provenance, staleness, Python readiness,
already loaded `libgdal` modules, legacy profile hooks, and a foreign
`gdalraster` package when `library(gdalraster)` would load a different tree.

## `library(gdalraster)` cannot find the package

Rebuild into the regular user library, or load an isolated install from its
library:

```{r}
#| eval: false
gdalraster.windows::gdal_setup()
library(gdalraster)
```

## `LoadLibrary` reports a missing module

A self-contained build must contain `gdalraster.dll`, `libgdal-*.dll`, and the
full dependency closure in the same `libs/x64` directory:

```{r}
#| eval: false
package_dir <- find.package("gdalraster")
list.files(file.path(package_dir, "libs", "x64"), pattern = "\\.dll$")
```

Run `gdal_update()` to rebuild and re-vendor from a complete bundle. Do not add
Rtools or another GDAL to `PATH` as a repair; that defeats package isolation.

## Empty Algorithm API or missing drivers

`gdal_verify()` requires:

- a non-empty `gdal_global_reg_names()` result;
- Arrow, Parquet, HDF5, and netCDF drivers;
- GEOS and EPSG:4326 resolution; and
- a successful first Parquet open in a fresh process.

If it fails, compare the runtime and build tags in `gdal_sitrep()`. A mismatch
requires `gdal_update()`. A matching build with missing drivers indicates a
bundle regression and should be reported with the verification output.

## Another GDAL is installed

GDAL executables from pixi, conda, OSGeo4W, Rtools, or MSYS2 may remain on the
machine. The package's `libs/x64` directory controls dependency resolution
while `gdalraster.dll` loads.

An identically named `libgdal` already loaded in the process is different:
Windows can reuse it by module name. Restart R, load `gdalraster` first, and
inspect `gdal_sitrep()` for the loaded module path.

## Legacy `.Rprofile` hooks

Version 0.4.0 removes activation and profile-hook APIs. Delete blocks marked:

```text
# >>> gdalraster.windows hook >>>
...
# <<< gdalraster.windows hook <<<
```

Also remove personal code that calls `load_gdal_dll()`,
`activate_gdal_runtime()`, or edits GDAL-related environment variables.
Rebuild once with `gdal_setup()` after cleanup.

## Locked runtime directory

Windows cannot delete mapped DLLs. `gdal_install_runtime(force = TRUE)` first
tries normal deletion, then moves remaining files into a sibling directory
named `<gdal_home>.stale-<pid>-<timestamp>`. Later installs remove stale
directories after locks disappear.

If move-aside also fails, another process holds the file without delete
sharing. Close other R sessions and File Explorer previews. PowerToys File
Locksmith or Resource Monitor can identify the holder.

## Locked `gdalraster` package

The builder stages the complete replacement before touching the installed
package. If the destination DLL is locked, replacement aborts and leaves the
old package unchanged. Restart R, ensure `gdalraster` is not loaded in another
session, then rerun the build.

## Embedded Python is unavailable

`gdal_sitrep()` reports whether the managed
`gdalraster-windows-osgeo-utils.pth` points at the current vendored
`gdalraster/python` directory.

For a regular user-library build use:

```{r}
#| eval: false
gdalraster.windows::gdal_enable_python()
```

When no system CPython is available, core GDAL functionality remains ready,
but Python-backed algorithms such as `driver gpkg validate` cannot run.

## Offline and proxy-restricted installation

Transfer a published bundle zip and pass it directly:

```{r}
#| eval: false
gdalraster.windows::gdal_setup(
  local_zip = "C:/Downloads/gdal-bundle-v3.13.2-windows-x64.zip"
)
```

The installer preserves an embedded manifest when present and otherwise
stamps provenance from the requested release information.
